Research Notes

### 2018-04-07 writups/notes.tex
check in agree_plots
write up some scratch work

For ease of computation, let's pick $F(x, t) = \cos \p*{\frac{2 \pi x}{L} -
\omega_0 t}$, so our full expected solution is (note $A + \epsilon =
Ae^{\epsilon/A} + \mathcal{O}(\epsilon^2)$)
\begin{equation}
    \begin{split}
        u_{1z}\p*{x, z, t \middle | \vec{k}, \omega}
            &= e^{z/2H}\frac{\cos\p*{\frac{2\pi x}{L} - \omega_0 t}
                \sin \p*{k_z(L - z)}}{\sin k_zL},\\
        u_{1x}\p*{x, z, t \middle | \vec{k}, \omega}
            &\approx -\frac{k_z}{k_x} e^{z/2H}\frac{
                \cos\p*{\frac{2\pi x}{L} - \omega_0 t + \frac{1}{2Hk_z}}
                \sin \p*{k_z(L - z)}
            }{\sin k_zL},\\
        \rho_1\p*{x, z, t \middle | \vec{k}, \omega}
            &\approx \frac{\rho_0}{H\omega} e^{z/2H}\frac{
                -\sin\p*{\frac{2\pi x}{L} - \omega_0 t}
                \sin \p*{k_z(L - z)}
            }{\sin k_zL},\\
        P_1\p*{x, z, t \middle | \vec{k}, \omega}
            &\approx -\frac{\rho_0 \omega k_z}{k_x^2} e^{z/2H}\frac{
                -\cos\p*{\frac{2\pi x}{L} - \omega_0 t + \frac{1}{2Hk_z}}
                \sin \p*{k_z(L - z)}
            }{\sin k_zL},\\
    \end{split}\label{eq:strat.part_sol}
\end{equation}
where $k_z: \omega(k_x, k_z) = \omega_0$ by the dispersion relation
\autoref{eq:strat_disp}. Note that $H$ contributes both to the overall
exponential profile and to the phase lag of $u_{1x}$.

@@

\subsection{Waveform Properties, Analytical}

Let's figure out the analytical forms for the phase, group velocity and the
energy density/power flux. We will try to recover these from waveforms.

@@

For our stratified atmosphere problem, $\omega^2 = \frac{N^2 k_x^2}{k^2 +
\frac{1}{4H^2}}$. Plugging in our parameters $k_z = -k_x = 2pi, N^2 = 10$ we can
find $\omega \approx 2.23\;\mathrm{Hz}$, so $\vec{v}_{ph} \approx
0.1777\p*{-\hat{x} + \hat{z}}\;\mathrm{s^{-1}}$.

Since we use a height of $5$, we should expect the initial disturbance to reach
the top boundary at $t \sim 30\;\mathrm{s}$.

On the other hand the group velocity is given $\vec{v}_g =
\pd{\omega}{k_i}\hat{\imath}$. In our problem, $v_{g, z} =
-\frac{N^2k_x^2k_z}{\p*{k_x^2 + k_z^2 + 1/4H^2}^{3/2}}$ while $v_{g, x} =
\frac{2N^2k_x}{\sqrt{k_x^2 + k_z^2 + 1/4H^2}} - \frac{N^2k_x^3}{\p*{k_x^2 +
k_z^2 + 1/4H^2}^{3/2}}$.

We should note that the kinetic energy density is just $E = \frac{1}{2}\rho
v^2$, the kinetic energy. Using then $\rd{E}{t} = 0$ shows that the
corresponding power flux is $\vec{P} = \frac{\vec{v}}{2}\rho v^2$.

\subsection{Extracting Information from the Waveform}

Let's make sure we can extract information from the linearized problem. Notable
things we should be interested in: phase/group velocity, energy density and
energy flux.

Based on our phase velocity expression \autoref{eq:v_ph}, if we have some
waveform $f(x, z, t)$, we can estimate its phase velocity $\vec{v}_{ph}(x, z,
t)$ by
\begin{equation}
    \vec{v}_{ph}(x, z, t) = \frac{\partial_t f
        \vec{\nabla}f}{\abs{\vec{\nabla}f}^2}.\label{eq:v_ph}
\end{equation}
This can be validated in the $f(x, z, t) \propto e^{i(\vec{k} \cdot \vec{x} -
\omega t)}$ case to give $\vec{v}_{ph} = \frac{\omega \vec{k}}{k^2}$. Note the
sign in \autoref{eq:v_ph}

@@

This requires we compute a ton of extra points though. Ideally, we want to use
BCs. TODO

\section{(Algebra/Optional) The Anelastic/Boussinesq Approximations}

>>>>

For ease of computation, let's pick $F(x, t) = \cos \p*{\frac{2 \pi x}{L} -
\omega_0 t}$, so our full expected solution is (note $A + \epsilon =
Ae^{\epsilon/A} + \mathcal{O}(\epsilon^2)$)
\begin{equation}
    \begin{split}
        u_{1z}\p*{x, z, t \middle | \vec{k}, \omega}
            &= e^{z/2H}\frac{\cos\p*{\frac{2\pi x}{L} - \omega_0 t}
                \sin \p*{k_z(L - z)}}{\sin k_zL},\\
        u_{1x}\p*{x, z, t \middle | \vec{k}, \omega}
            &\approx \frac{k_z}{k_x} e^{z/2H}\frac{
                \cos\p*{\frac{2\pi x}{L} - \omega_0 t + \frac{1}{2Hk_z}}
                \sin \p*{k_z(L - z)}
            }{\sin k_zL},\\
        \rho_1\p*{x, z, t \middle | \vec{k}, \omega}
            &\approx \frac{\rho_0}{H\omega} e^{z/2H}\frac{
                -\sin\p*{\frac{2\pi x}{L} - \omega_0 t}
                \sin \p*{k_z(L - z)}
            }{\sin k_zL},\\
        P_1\p*{x, z, t \middle | \vec{k}, \omega}
            &\approx -\frac{\rho_0 \omega k_z}{k_x^2} e^{z/2H}\frac{
                \cos\p*{\frac{2\pi x}{L} - \omega_0 t + \frac{1}{2Hk_z}}
                \sin \p*{k_z(L - z)}
            }{\sin k_zL},\\
    \end{split}\label{eq:strat.part_sol}
\end{equation}
where $k_z: \omega(k_x, k_z) = \omega_0$ by the dispersion relation
\autoref{eq:strat_disp}. Note that $H$ contributes both to the overall
exponential profile and to the phase lag of $u_{1x}$.

@@

\subsection{Phase/Group Velocity}

Let's figure out the analytical forms for the phase, group velocity and the
energy density/power flux.

@@

For our stratified atmosphere problem, $\omega^2 = \frac{N^2 k_x^2}{k^2 +
\frac{1}{4H^2}}$. This corresponds to $\vec{v}_{ph} = \frac{Nk_x}{\sqrt{k_x^2 +
k_z^2 + 1/4H^2}} \vec{k}$.

On the other hand the group velocity is given $\vec{v}_g =
\pd{\omega}{k_i}\hat{\imath}$. In our problem, $v_{g, z} =
-\frac{Nk_xk_z}{\p*{k_x^2 + k_z^2 + 1/4H^2}^{3/2}}$ while $v_{g, x} =
\frac{N}{\sqrt{k_x^2 + k_z^2 + 1/4H^2}} - \frac{Nk_x^2}{\p*{k_x^2 +
k_z^2 + 1/4H^2}^{3/2}} = \frac{N\p*{k_z^2 + 1/4H^2}}{\p*{k_x^2 +
k_z^2 + 1/4H^2}^{3/2}}$.

\subsection{Energy/Power Flux}

To compute the energy and power flux of the wave, we recall that for a general
fluid the energy conservation equation reads
\begin{equation}
    \pd{}{t}\p*{\frac{1}{2}\rho v^2 + \rho \epsilon}
        = \vec{\nabla} \cdot \p*{
            \rho \vec{v}\p*{v^2 + \epsilon + \frac{P}{\rho}}
        },
\end{equation}
where $\epsilon$ is the internal energy. But since $P = (\gamma -
1)\rho\epsilon$ and we take $\gamma \to \infty$ incompressible limit, $\epsilon
= 0$, so the energy of the wave is just $\frac{1}{2}\p*{\rho_0 + \rho_1} u_1^2$
and the power flux is $\p*{\rho_0 + \rho_1}\vec{u}_1u_1^2 + P_1\vec{u}_1$.

Another formula given by Sutherland 2011 is $\pd{\expvalue{E}}{t} = v_{g, z}
\expvalue{E}$ where we average over $x$ one wavelength. I wasn't able to show
that these two agree in general; Sutherland's formula seems to be derived for a
traveling wavepacket whereas we have no such thing. We do not use this
expression since the Landau \& Lifshitz expression works well.

@@

In order to compute the reflecting boundary condition, we must find a
sequence of differential operators that well approxmiates the system of PDEs. We
can get this from the dispersion relation; since our radiating boundary is along
$z$, we should try to solve for $k_z$ in the dispersion relation to get a
pseudodifferential operator form for $\pd{}{z}u_z$ (we take $k_xH \ll 1$ as well
to simplify):
\begin{align}
    k_z^2 &= \frac{N^2}{\omega^2}k_x^2 - k_x^2 - \frac{1}{4H^2},\nonumber\\
        &\approx k_x^2\p*{\frac{N^2}{\omega^2} - 1},\\
    \pd{u_z}{z} &\approx -\pd{u_z}{x}\sqrt{\frac{N^2}{\omega^2} - 1}.
\end{align}
We pick the negative sign in accordance with an outgoing wave, per the group
velocity formulae for $k_z > k_x$.

\clearpage

\appendix

\chapter{Deriving Fluids Results}

\section{Equation of Energy Conservation}

We follow Landau \& Lifshitz's derivation for this expression. Consider that the
total energy stored in a wave must be the sum of its kinetic and internal energy
$\frac{1}{2}\rho v^2 + \rho \epsilon$ where $\epsilon$ is the internal energy
density. To obtain an equation of energy conservation, we must take the time
derivative of this expression. First, we consider
\begin{align}
    \pd{}{t}\p*{\frac{1}{2}\rho v^2} &= \frac{1}{2}v^2\pd{\rho}{t}
        + \rho \vec{v} \cdot \pd{\vec{v}}{t},\nonumber\\
        &= -\frac{1}{2}v^2 \p*{\vec{\nabla} \cdot \rho \vec{v}}
            - \rho \vec{v} \cdot \p*{
                \p*{\vec{v} \cdot \vec{\nabla}}\vec{v} +
                    \frac{\vec{\nabla}P}{\rho} - \vec{g}
            }_{T, V},\nonumber\\
        &= -\frac{1}{2} v^2 \p*{\vec{\nabla} \cdot \rho \vec{v}}
            - \frac{1}{2}\rho \vec{v} \cdot \p*{\vec{\nabla} v^2}
            - \vec{v} \cdot \p*{
                \rho \vec{\nabla} w - \rho T \vec{\nabla}s
            } + \rho \vec{v} \cdot \vec{g}.
\end{align}
We have denoted $s$ the specific internal entropy density of the fluid and
$\mathrm{d}w = T\mathrm{d}s + \frac{\mathrm{d}P}{\rho} = T\mathrm{d}s +
\epsilon$ the specific internal enthalpy density of the fluid.
Recall enthalpy $\epsilon = w - Ts$ is the usual thermodynamic definition.

At the same time, consider
\begin{align}
    \pd{}{t}\p*{\rho \epsilon} &= \epsilon \pd{\rho}{t}
        + \rho \pd{}{t}\p*{Ts - \frac{P}{\rho}}_{T, P},\nonumber\\
        &= \epsilon \pd{\rho}{t} + \rho T \pd{s}{t} + \frac{P}{\rho}\pd{\rho}{t}
            ,\nonumber\\
        &= w \pd{\rho}{t} + \rho T \pd{s}{t},\nonumber\\
        &= -w\vec{\nabla}\p*{\rho \vec{v}} - \rho T \vec{v} \cdot \p*{
            \vec{\nabla}s
        }.
\end{align}
Summing the two, we find
\begin{equation}
    \pd{}{t}\p*{\frac{\rho v^2}{2} + \rho \epsilon}
        = -\vec{\nabla} \cdot \s*{
            \rho \vec{v}\p*{v^2 + w}
        } = -\vec{\nabla} \cdot \s*{
            \rho \vec{v}\p*{v^2 + \epsilon + \frac{P}{\rho}}
        }.
\end{equation}

\section{The Anelastic/Boussinesq Approximations}
